---
title: "Performing differential gene expression analysis"
author: "Friederike Duendar, ABC"
output:
  html_document: default
  pdf_document: default
  toc: true
editor_options: 
  chunk_output_type: console
---

```{r settingUp, warning=FALSE, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE,cache=TRUE)
library(DESeq2)
library(magrittr)
load("~/Documents/Teaching/2018_RNA-seq/fall_2018/Rclass.RData")
```

We need to ensure that the fold change will be calculated using the WT as the base line.
`DESeq` used the levels of the condition to determine the order of the comparison.

```{r DESeqPrep_relevel}
str(DESeq.ds$condition)
DESeq.ds$condition <- relevel(DESeq.ds$condition, ref="WT")
str(DESeq.ds$condition)
```

#### Analysis design

```{r}
design(DESeq.ds)
```

### Running the DE analysis

```{r}
DESeq.ds <- DESeq(DESeq.ds)
```

This one line of code is equivalent to these three lines of code:

```{r eval=FALSE}
DESeq.ds <- estimateSizeFactors(DESeq.ds) # sequencing depth normalization between the samples
DESeq.ds <- estimateDispersions(DESeq.ds) # gene-wise dispersion estimates across all samples
DESeq.ds <- nbinomWaldTest(DESeq.ds) # this fits a negative binomial GLM and applies Wald statistics to each gene
```

Extract the base means across samples, log2 fold changes, standard errors, 
test statistics, p-values and adjusted p-values for every gene using `results()`.

```{r}
resultsNames(DESeq.ds) # tells you which types of values can be extracted with results()
DGE.results <- results(DESeq.ds,
                       independentFiltering = TRUE,
                       alpha = 0.05)
head(DGE.results) # the first line will tell you which comparison was done to achieve the log2FC
summary(DGE.results)

# the DESeqResult object can basically be handled like a data.frame
table(DGE.results$padj < 0.05)
```

NAs in the `padj` column (but values in both `log2FC` and `pvalue`) are indicative of that gene being filtered out by the independent filtering [because it was very lowly expressed].

The __MA-plot__ provides a global view of the differential genes, with the log2 fold
change on the y-axis over the mean of normalized counts.

Genes that pass the significance threshold (adjusted p.value \textless 0.05) are colored in red.

```{r MAplots}
plotMA(DGE.results, alpha = 0.05, 
       main = "Test: p.adj.value < 0.05", ylim = c(-4,4))
```

A adj. p-value histogram:

```{r adjpvalueHistogram}
hist(DGE.results$padj, 
     col="grey", border="white", xlab="", ylab="", main="frequencies of adj. p-values\n(all genes)")
```

A sorted results table:

```{r}
DGE.results.sorted <- DGE.results[order(DGE.results$padj),]
head(DGE.results.sorted)
```

Plotting counts for single genes (seq. depth normalized, log2-transformed)

```{r singleGenePlot}
par(mfrow=c(1,2))
plotCounts(DESeq.ds, gene="YAL056W", normalized = TRUE)  
plotCounts(DESeq.ds, gene=which.max(DGE.results$padj), main = "Max. p.adj.")
```

`plotCounts` simply uses `counts(dds, normalized = TRUE) + 0.5`.

You can also use `pcaExplorer` for individual gene plots of `rlog` values.

A heatmap of the genes that show differential expression with adjusted p-value \textless 0.05 :

```{r heatmaps}
# identify genes with the desired adjusted p-value cut-off
DGEgenes <- rownames(subset(DGE.results.sorted, padj < 0.05))

# extract rlog-transformed values into a matrix
rlog.dge <- DESeq.rlog[DGEgenes,] %>% assay

library(pheatmap)
# heatmap of DEG sorted by p.adjust
pheatmap(hm.mat_DGEgenes, scale="none", show_rownames = FALSE, main = "DGE (no scaling)")
pheatmap(hm.mat_DGEgenes, scale="row", show_rownames = FALSE, main = "DGE (row-based z-score)")
```


## Number 1 sanity check: is SNF2 affected in the SNF2 mutant yeast samples?

To find this out, we need to retrieve the gene names and match them to the ORF 
IDs that we've used so far.
[http://www.bioconductor.org/packages/3.1/data/annotation/](http://www.bioconductor.org/packages/3.1/data/annotation/)
lists annotation packages that are available within R through bioconductor.

We will go with `org.Sc.sgd.db`.

```{r checkForSNF2}
#source("http://bioconductor.org/biocLite.R")
#biocLite("org.Sc.sgd.db")
library(org.Sc.sgd.db) # org.Hs.eg.db, org.Mm.eg.db

# list keytypes that are available to query the annotation data base
keytypes(org.Sc.sgd.db)

# list columns that can be retrieved from the annotation data base
columns(org.Sc.sgd.db)

# make a batch retrieval for all DE genes
DGEgenes <- rownames(subset(DGE.results.sorted, padj < 0.05))

anno.DGE <- select(org.Sc.sgd.db,
               keys = DGEgenes, # rownames
               keytype="ORF", # our rownames are ORF identifiers
               columns=c("SGD","GENENAME")) # what to return

# check whether SNF2 pops up among the top downregulated genes
#head(anno.DGE[match(DGEgenes, anno.DGE$ORF),])
head(anno.DGE)

```

To get a feeling for how the difference between WT and snf2 ko looks like for a
housekeeping gene, let's repeat the exercise.

```{r singleGenePlot_noDGE, echo = TRUE}
par(mfrow=c(1,2))
plotCounts(dds = DESeq.ds,
           gene = "YOR290C",
           normalized = TRUE, transform = FALSE,
           main = expression(atop("Expression of "*italic("snf2"), "(YOR290C)")))

plotCounts(dds = DESeq.ds,
           gene = "YGL012W", # the last gene in DGE
           normalized = TRUE, transform = FALSE,
           main = expression(atop("Expression of "*italic("erg4"), "(YGL012W)")))

```

Export the log2FC, p-values etc. into a text file:

```{r, eval=FALSE}
out.df <- merge(as.data.frame(DGE.results), anno.DGE, by.x = "row.names", by.y = "ORF")

write.table(subset(out.df, padj < 0.05), file = "DESeq2results_WT-vs-SNF2.txt",
            sep = "\t", quote = FALSE, row.names = FALSE)
```

-----------------------

# What next?

## GO term enrichment

*Among our list of DE genes, which GO terms are enriched?*

Transcripts that are longer or more highly expressed give more statistical power
for detecting differential expression between samples

The same bias holds true for GO categories: categories with predominantly highly
expressed or long genes are more likely to be found to be over-represented within
the DEG.

`GOseq`:

1. determine DEG
2. quantify likelihood of DE as a function of gene length (--> weight)
3. statistical test of each GO category's significance taking the DE probability into account

Couldn't get it to work with yeast.

```{r goseq_yeast, eval = FALSE}
# extract Entrez IDs for the genes of interest
##DGEgenes.entrez <- subset(anno.all, ORF %in% DGEgenes, select = "ENTREZID")[,1] 
##limma::goana(DGEgenes.entrez, geneid = DGEgenes.entrez)
#DGEgenes.GO <- select(org.Sc.sgd.db, keys = DGEgenes, keytype = "ORF", columns = c("ORF", "GENENAME", "GO"))
```

Manual work-around for mouse:

```{r goseq_mm}
#biocLite("org.Mm.eg.db")
#biocLite("goseq")
#biocLite("biomaRt")
library(biomaRt)
library(org.Mm.eg.db)
library(goseq)
library(geneLenDataBase)

# gene list from an experiment involving immune cells
gns <- read.table("~/Documents/Teaching/2018_RNA-seq/fall_2018/mm_genes.txt", stringsAsFactors = FALSE, skip = 1)

# retrieve the ENSEMBL symbols for the gene names
anno.mm <- select(org.Mm.eg.db,
               keys = gns$V1, 
               keytype="SYMBOL", 
               columns=c("ENSEMBL","SYMBOL","GENENAME")) # what to return

# in addition, retrieve all possible mouse gene names
# because goseq wants to know about the universe of *all* genes
mart <- useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")
mm_all <- getBM(attributes = c("ensembl_gene_id"),  mart = mart)

# get length data using goseq's data base of gene lengths
# first, load the data from the geneLenDataBase package
data(list = "mm9.ensGene.LENGTH", package = "geneLenDataBase")

# split the length values per ENSEMBL ID into a list
len_data = split(mm9.ensGene.LENGTH$Length, 
            mm9.ensGene.LENGTH$Gene)

# get the gene names from the length data base
gn_names = mm9.ensGene.LENGTH$Gene

# calculate the median (and other) lengths summary values for the 
# different genes
len = data.frame(Gene = names(len_data), 
                 Median = sapply(len_data,median), 
                 Min = sapply(len_data, min),
                 Max = sapply(len_data, max),
                 Count = sapply(len_data, length))

# goseq wants a named binary vector where 
# 1  = DE, 0 = not DE 
# and the names of the [0,1]-vector are gene IDs
gogns <- len$Gene %in% anno.mm$ENSEMBL %>% as.integer
names(gogns) <- len$Gene

# Quantifying the lenght bias (= weight for each gene)
# using a Probability Weighting Function (PWF): probability of a gene being DE ~ length
# proportion of DE genes is plotted as a function of the transcript length
pwf <- goseq::nullp(gogns, bias.data = len$Median) 

# do the actual test for enrichment of GO terms
GO.wall <- goseq(pwf, "mm9", "ensGene")
head(GO.wall)

# can be summarized in http://revigo.irb.hr/
subset(GO.wall, over_represented_pvalue < 0.01,select = c("category","over_represented_pvalue")) %>%
  write.table(., file = "~/Documents/Teaching/2018_RNA-seq/fall_2018/GOterms.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

